NanoMaLy

### 2015-11-13 paper/pldi16-submission/semantics.tex
misc
search loop

  Let $s_k = \ptype{k}{f}$. Suppose \hastype{v}{\tincompat{t}{s_{k+1}}}, we
  will show that \stepi{k+1}{\eapp{f}{v}}{\emptysu}{\stuck}{\su}.
  \begin{description}
  \item[Case \tincompat{t}{s_k}:] 
    The inductive hypothesis applies.
  \item[Case \tcompat{t}{s_k} but \tincompat{t}{s_{k+1}}:] 
    By Lemma~\ref{lem:force-inst} we must have called \forcesym at step
    $k+1$. 
    %
    A case analysis of the applicable rules shows that \forcesym cannot
    have succeeded.
  \end{description}
\end{proof}
%
Now we can prove our soundness theorem.
%
\begin{thm}
\label{thm:soundness}
  For any function $f$, if \steps{\eapp{f}{\ehole{}}}{\emptysu}{\stuck}{\su},
  then $\forall t. \exists \hastype{v}{t}. \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}$.
\end{thm}
\begin{proof}
  Let $\tfun{s}{\_} = \ptype{}{f}$.
  \begin{description}
  \item[Case \tcompat{s}{t}:] Our witness is already valid.
  \item[Case \tincompat{s}{t}:] We can apply Lemma~\ref{lem:k-stuck} and
    use \emph{any} \hastype{v}{t} as a witness for $f$.
  \end{description}
\end{proof}

@@

>>>>

  Let $\tfun{s_k}{\_} = \ptype{k}{f}$. Suppose \hastype{v}{\tincompat{t}{s_{k+1}}}, we
  will show that \stepi{k+1}{\eapp{f}{v}}{\emptysu}{\stuck}{\su}.
  \begin{description}
  \item[Case \tincompat{t}{s_k}:] 
    The inductive hypothesis applies.
  \item[Case \tcompat{t}{s_k} but \tincompat{t}{s_{k+1}}:] 
    By Lemma~\ref{lem:force-inst} we must have called \forcesym at step
    $k+1$. 
    %
    A case analysis of the applicable rules shows that \forcesym cannot
    have succeeded.
  \end{description}
\end{proof}
%
Now we can prove our soundness theorem.
%
\begin{thm}
\label{thm:soundness}
  For any function $f$, if \steps{\eapp{f}{\ehole{}}}{\emptysu}{\stuck}{\su},
  then $\forall t. \exists \hastype{v}{t}. \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}$.
\end{thm}
\begin{proof}
  Let $\tfun{s}{\_} = \ptype{}{f}$.
  \begin{description}
  \item[Case \tcompat{s}{t}:] Our witness is already valid.
  \item[Case \tincompat{s}{t}:] We can apply Lemma~\ref{lem:k-stuck} and
    use \emph{any} \hastype{v}{t} as a witness for $f$.
  \end{description}
\end{proof}

@@

The top-level search loop takes as input an open program -- a sequence
of binders -- and searches for an expression that closes the input
program and gets stuck. Concretely, given an input
%
\begin{code}
  let f1 = e1 in
  let f2 = e2 in
  ...
  let fn = en in
\end{code}
%
we will search for an expression of the form:
%
\begin{code}
  fn v1 ... vn
\end{code}
%
Figure~\ref{fig:expression-api} describes a small API for manipulating
and evaluating \lang expressions, which we will use to define our search
algorithm.
%
The bulk of the search is performed by @eval@, which
implements our operational semantics from \S\ref{sec:semantics}.
%
The operational semantics is non-deterministic due to \gensym,
thus @eval@ returns a list of possible results.
%
\begin{figure}[t]
  \centering
  \begin{mcode}
  -- transitive small-step reduction, 
  -- returning a list of results
  eval :: ($e$, $\su$) -> [($v$, $\su$)]

  -- manipulating expressions
  subst   :: $\su$ -> $v$ -> $v$
  mkApps  :: $e$ -> [$e$] -> $e$
  mkLets  :: [($x$, $e$)] -> $e$ -> $e$
  isStuck :: $v$ -> Bool
  \end{mcode}
  \caption{Expression API}
  \label{fig:expression-api}
\end{figure}
%
We also define a few helper functions for manipulating expressions:
\begin{itemize}
\item @subst@ applies a substitution of holes to a value,
\item @mkApps@ creates a nested sequence of applications in the usual
  left-associative style,
\item @mkLets@ takes a list of binders and a body expression, and
  creates a sequence of nested let-binders, and
\item @isStuck@ tests whether a value is the \stuck term.
\end{itemize}
%
\begin{figure*}[t]
  \centering
  \begin{mcode}
  check :: [($x$, $e$)] -> Result
  check bnds =
    -- (2) search for a witness
    case find (isStuck . fst) results of
      Nothing      -> Safe
      Just (_, su) -> Unsafe (mkApps f (subst su args))
    where
      (args, results) = loop []
      f               = snd (last bnds)
      build args      = mkLets bnds (mkApps f args)

      -- (1) find the correct number of arguments
      loop :: [$v$] -> ([$v$], [($v$, $\su$)])
      loop args = case eval (build args, []) of 
        ($\efun{x}{e}$, _) : _ -> loop (args `snoc` $\ehole{}$)
        results      -> (args, results)  
  \end{mcode}
  \caption{Implementing our counter-example search in terms of the
    operational semantics. 
    \ES{1 and 2 are backwards, ugh...}
    \ES{should address case where output types of successive runs dont match}
  }
  \label{fig:search-algo}
\end{figure*}
%
Figure~\ref{fig:search-algo} summarizes the overall implementation of
our search for witnesses, which takes as input a sequences of binders
and returns either @Safe@ if no witness could be found, or |Unsafe $e$|
where $e$ is a term that causes the input program to get stuck.
